Stochastic Blockmodels for Relational Event Data

### 2012-04-27 docs/submission.tex
Update split merge algorithm to new idea.

Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  One class of statistical models for network data uses latent variables to help model unobserved heterogeneity.  For example, the stochastic blockmodel \cite{Nowicki2001, Kemp, Ishiguro2010} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between nodes of each block.  % Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.

@@

This is a two-stage MH procedure.  First, pick two nodes $i,j$ uniformly.  If they belong to the same block $k$ (i.e. $z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ with new cluster $l$:
\begin{itemize}
\item For blocks $r \ne l$, sample $\phi_{lr}^{split} \sim N(\phi_{kr}^{merge},\sigma^2)$ and $\phi_{rl}^{split} \sim N(\phi_{rk}^{merge},\sigma^2)$ and $\phi_{ll}^{split} \sim N(\phi_{kk}^{merge}, \sigma^2)$.
\item For all $a$ such that $z_a^{merge} = k$, randomly initialize $z_a^{split}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{split}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{split}_{-a},\boldsymbol{\phi}^{split})$
where $n_k=\sum_{i}I(z_i^{split}=k)$.% and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
\end{itemize}

 If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e. $z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
\begin{itemize}
\item For all blocks $r \ne k,l$: sample $\phi_{kr}^{merge} \sim \mbox{Unif}(\{\phi_{kr}^{split}, \phi_{lr}^{split}\})$, and $\phi_{rk}^{merge} \sim \mbox{Unif}(\{\phi_{rk}^{split}, \phi_{rl}^{split}\})$.%, and sample $\phi_{lr}^{merge},\phi_{rl}^{merge}$, and $\phi_{ll}^{merge}$ independently from the prior.
\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
\end{itemize}

 The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(Y|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$

The first term is the likelihood ratio of the two states.  Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{split}$, the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan from  $\mathbf{z}^{merge}$ is computed as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}^{merge}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{split})$.  Since there is only one way of combining two clusters we know $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$.   The transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ is a product of proposal densities and
$q(\boldsymbol{\phi}^{split} \rightarrow
\boldsymbol{\phi}^{merge}) = \frac{1}{2}\prod_{r \ne k,l}\left(\frac{1}{2}\right)^2$. \footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{merge})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.

 \appendix{\textbf{Split-merge Metropolis-Hastings Move for Relational Event Blockmodel}}

This is a two-stage MH procedure, first picking two nodes at random then creating a new set of parameters.  We also use an auxilary variable to avoid changing dimensionality (similar to Neal 2000, alg. 8).

Next pick two nodes $i,j$ uniformly.  If they belong to the same block $k$ (i.e. $z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
\begin{itemize}
%\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
\item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
%\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
\item Let $S$ be all nodes assigned to $k$.  Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
\item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
\end{itemize}

  If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e. $z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
\begin{itemize}
\item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$, and sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
%\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
\item Let $S$ be the set of nodes assigned to $k$ or $l$. Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
\item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
\end{itemize}

The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$

For the proposed merge step,
$$q(\boldsymbol{\phi}^{split} \rightarrow
\boldsymbol{\phi}^{merge}) = \prod_{q,r \ne l}\frac{1}{2} \prod_r p(\phi_{r,l,p}|\mu_p,\sigma_p^2)p(\phi_{l,r,p}|\mu_p,\sigma_p^2) p(\phi_{l,l,p}|\mu_p,\sigma_p^2)$$ though the latter product cancels with a corresponding term in $p(\boldsymbol{\phi}^{merge})$.  We can compute the transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.  For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.  $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.

>>>>

Statistical methods for analyzing network data have become increasingly useful for studying  phenomenon ranging from people communicating online to protein interactions \cite{Goldenberg2009}.  One class of statistical models for network data uses latent variables to help model unobserved heterogeneity.  For example, the stochastic blockmodel \cite{Nowicki2001, Kemp, Ishiguro2010} assumes each node in the network belongs to some block (or cluster) and parametrizes the probability of edges between nodes of each block.  % Such approaches are especially useful for large-scale network data where higher-order dependencies can cause models such as exponential random graph models to be too complex to fit.
% TODO: Cite Rodriguez: Modeling dynamics of social networks via h. blockmodels

@@

This is a two-stage MH procedure.  First, pick two nodes $i,j$ uniformly.  If they belong to the same block $k$ (i.e. $z_i=z_j=k$), then let the ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ with new cluster $l$:
\begin{itemize}
\item Sample $\phi_{kk}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and $\phi_{ll}^{s} \sim N(\phi_{kk}^{m},\sigma^2)$ and for blocks $r \ne l$, sample $\phi_{lr}^{s} \sim N(\phi_{kr}^{m},\sigma^2)$ and $\phi_{rl}^{s} \sim N(\phi_{rk}^{m},\sigma^2)$ 
\item For all $a$ such that $z_a^{m} = k$, randomly initialize $z_a^{s}$ to either $k$ or $l$, then perform a restricted Gibbs scan by sampling $p(z_{a}^{s}=k|\cdot)  \propto n_k p(A|\mathbf{z}^{s}_{-a},\boldsymbol{\phi}^{s})$
where $n_k=\sum_{i}I(z_i^{s}=k)$.% and compute $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})$ to be the product of the probabilities for the sampled assignments.
\item Accept  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s})$ as the new state with probability $\alpha^*$, defined below.
\end{itemize}

 If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e. $z_i = k \ne z_j=l$), we set the ``split'' state  $(\boldsymbol{\phi}^{s},\mathbf{z}^{s}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as follows:
\begin{itemize}
\item Sample $\phi_{kk}^m \sim N((\phi_{kk}^s + \phi_{ll}^s)/2,
  \sigma^2)$.  Sample $\phi_{ll}^m \sim N(0,\tau^2)$ and for all blocks $r \ne k,l$ sample $\phi_{kr}^{m} \sim N(0,\tau^2)$ (the prior)
\item For all $a$ such that $z_a \in \{k,l\}$ set $z_a^m = k$.
\item Accept $(\boldsymbol{\phi}^{m},\mathbf{z}^{m})$ as the new state with probability $1/\alpha^*$, defined below.
\end{itemize}

 The Metropolis-Hastings acceptance probabilities require us to compute 
$$\alpha^* =\frac{p(Y|\boldsymbol{\phi}^{s},\mathbf{z}^{s})}{p(Y|\boldsymbol{\phi}^{m},\mathbf{z}^{m})}  \frac{p(\boldsymbol{\phi}^{s})}{p(\boldsymbol{\phi}^{m})} \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} \frac{q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})}{q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s})} \frac{q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})}{q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})}$$

The first term is the likelihood ratio of the two states.  Letting $S$ be the set of nodes assigned to $k$ or $l$ in $\mathbf{z}^{s}$, the probability of sampling $\mathbf{z}^{s}$ during a restricted Gibbs scan from  $\mathbf{z}^{m}$ is computed as $q( \mathbf{z}^{m} \rightarrow \mathbf{z}^{s}) = \prod_{s \in S} p(z_{s}=z_s^{s} |\{ z_a^{s},z_{b}^{m}: a  < s \ \mbox{and} \ s < b\}, \boldsymbol{\phi}^{s})$.  Since there is only one way of combining two clusters, we know $q(\mathbf{z}^{s} \rightarrow \mathbf{z}^{m})=1$.   The transition $q(\boldsymbol{\phi}^{m} \rightarrow \boldsymbol{\phi}^{s})$ and 
$q(\boldsymbol{\phi}^{s} \rightarrow
\boldsymbol{\phi}^{m})$ are products of Normal densities. \footnote{To ensure both states have the same dimensionality, one can optionally sample the parameters empty cluster $l$, though the extra term  in $q(\boldsymbol{\phi}^{s} \rightarrow \boldsymbol{\phi}^{m})$ would cancel with a corresponding term in $p(\boldsymbol{\phi}^{m})$}  Finally, under this prior $ \frac{p(\mathbf{z}^{s})}{p(\mathbf{z}^{m})} = \frac{(n_{z_i}^{s} - 1)!(n_{z_j}^{s} - 1)!}{(n_{z_i}^{m}-1)!}$.

%  \appendix{\textbf{Split-merge Metropolis-Hastings Move for Relational Event Blockmodel}}

% Next pick two nodes $i,j$ uniformly.  If they belong to the same block $k$ (i.e. $z_i=z_j=k$), then we call the current state the ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``split'' state $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ by letting $l=K+1$ and for $p \in [1,P], r \in [1,K]$
% \begin{itemize}
% %\item For $p \in [1,P], r \in [1,K]$: let $l=K+1$ sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$
% \item Sample $\phi_{lrp}^{split} \sim N(\phi_{krp}^{merge},\sigma^2)$ and $\phi_{rlp}^{split} \sim N(\phi_{rkp}^{merge},\sigma^2)$ and $\phi_{llp}^{split} \sim N(\phi_{kkp}^{merge}, \sigma^2)$.
% %\item Sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
% \item Let $S$ be all nodes assigned to $k$.  Randomly assign $z_a^{split}$ and perform restricted Gibbs scan for $\mathbf{z}^{split}$: $$p(z_{a}^{split}=k|\cdot)  \propto m_k p(A|\mathbf{z}^{split}_{-a},\phi^{split})$$
% where $m_k=\sum_{i \in S}I(z_i=k)$ and compute $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})$ to be the product of the probabilities for the sampled assignments.
% \item Accept  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split})$ as the new state with probability $\alpha^*$, defined below.
% \end{itemize}

%   If $i$ and $j$ belong to different blocks $k$ and $l$ (i.e. $z_i = k \ne z_j=l$), we define the current state to be the ``split'' state  $(\boldsymbol{\phi}^{split},\mathbf{z}^{split}) = (\boldsymbol{\phi},\mathbf{z})$ and propose a ``merge'' state $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as follows:
% \begin{itemize}
% \item For each $p \in [1,P]$ and for all blocks $r \in [1,K], r\ne l$ draw $\phi_{krp}^{merge}$ uniformly from $\{\phi_{krp}^{split}, \phi_{lrp}^{split}\}$,  $\phi_{rkp}^{merge}$ uniformly from $\{\phi_{rkp}^{split}, \phi_{rlp}^{split}\}$, and sample $\phi_{lrp}^{merge},\phi_{rlp}^{merge}$, and $\phi_{llp}^{merge}$ independently from the prior $N(\mu_p,\sigma^2_p)$.
% %\item Sample $\phi_{lrp}^{split},\phi_{rlp}^{split}$, and $\phi_{llp}^{split}$ independently from the prior $N(\mu_p,\sigma^2_p)$
% \item Let $S$ be the set of nodes assigned to $k$ or $l$. Let $\mathbf{z}$ be a random assignment for nodes a and compute the probability of sampling $\mathbf{z}^{split}$ during a restricted Gibbs scan as $q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split}) = \prod_{s \in S} p(z_{s}=z_s^{split} |\{ z_a^{split},z_{b}: a  < s, s < b\}, \boldsymbol{\phi}^{split})$
% \item Accept $(\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})$ as the new state with probability $1/\alpha^*$, defined below.
% \end{itemize}

% The Metropolis-Hastings acceptance probability for the split move is: $$\alpha^* =\frac{p(A|\boldsymbol{\phi}^{split},\mathbf{z}^{split})}{p(A|\boldsymbol{\phi}^{merge},\mathbf{z}^{merge})}  \frac{p(\boldsymbol{\phi}^{split})}{p(\boldsymbol{\phi}^{merge})} \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} \frac{q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})}{q( \mathbf{z}^{merge} \rightarrow \mathbf{z}^{split})} \frac{q(\boldsymbol{\phi}^{split} \rightarrow \boldsymbol{\phi}^{merge})}{q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})}$$

% For the proposed merge step,
% $$q(\boldsymbol{\phi}^{split} \rightarrow
% \boldsymbol{\phi}^{merge}) = \prod_{q,r \ne l}\frac{1}{2} \prod_r p(\phi_{r,l,p}|\mu_p,\sigma_p^2)p(\phi_{l,r,p}|\mu_p,\sigma_p^2) p(\phi_{l,l,p}|\mu_p,\sigma_p^2)$$ though the latter product cancels with a corresponding term in $p(\boldsymbol{\phi}^{merge})$.  We can compute the transition $q(\boldsymbol{\phi}^{merge} \rightarrow \boldsymbol{\phi}^{split})$ directly as a product of Normal densities.  For the block assignments, $q(\mathbf{z}^{split} \rightarrow \mathbf{z}^{merge})=1$ since  there is only one way of combining two clusters.  $ \frac{p(\mathbf{z}^{split})}{p(\mathbf{z}^{merge})} = \frac{(n_{z_i}^{split} - 1)!(n_{z_j}^{split} - 1)!}{(n_{z_i}^{merge}-1)!}$.
